{ \normalsize Master Thesis} \\ GEFILOC: Generic Framework for Indoor Location Systems

### 2017-04-15 paper.tex
Final V1

\title{{ \normalsize Master Thesis} \\
	Bluetooth Low Energy Supported Indoor Location}
\author{
	Ricardo Martins {\tt ricardo.pachecomartins@gmail.com}
	Instituto Superior T\'{e}cnico}
\date{\today}

@@

	Abstract

@@

Outdoor positioning systems have greatly evolved in the past with the global positioning system (GPS) as its most important system. Since GPS is an outdoor position systems and is based on a network of satelite, when the required scenario for position tracking is inside a building, new constraints are presented onto the process such as the attenuation and reflection of eletromagnetic waves upon collision with building walls and obstacles \cite{survey1}. As such there was a need to find reliable indoor systems that by nature would reduce the impact of some of the mentioned constraints.

In order to understand indoor position there is a need to understand the full scope of variables that comes to surface when moving from outdoor to indoor. When developing an indoor system there is a need to make sure that it can tackle the chalenges such as small space dimension which reforce the need for higher precision, a higher probability of non existent line of sight, influence of obstacles such as walls, furniture, moveable objects such as doors and human beings\cite{reviewtechniques, survey3}. All of the previously mentioned affect the way electromagnetic waves propagate in an indoor environment leading to problems related to severe multipath and reflection on existent surfaces \cite{surveywireless}. Besides propagation challenges, there are energy consumption, accuracy and deployment costs that play a critical role in deciding the viability of a proposed indoor location technique. Another interesting challenges that also surfaces is the means to describe a location. When dealing with outdoor location systems, the provided location is always characterized by two values, latitude and longitude, since the existant environment, i.e. the planet Earth, is always the same and as such the coordinates system can be relative to it. In indoor systems, the surrounding environment can be of many shapes and as such different ways of representing the data are necessary. If one thinks about providing indoor location on an office, the precise location isn't as relevant as just knowing the general location, i.e. a building specific description of the location in the form of building, floor and room. On other environments such as supermarkets or even in the previous example, where the objective is to provide more a precise location description, a cartesian coordinate system (x,y) is required.

With the evolution of mobile devices there has been a sizeable number of different technologies that can possibly be used in indoor locationing \cite{surveythesis,survey2} such as GPS-based technologies, using high sensitivity antenas to overcome GPS's indoor issues, RFID , Wireless LAN and Bluetooth among others, allowing even for hibrid systems which make use of more than one of the technologies mentioned above.

We are now in an era dominated by smartphones and as such they have became the central piece of several indoor location system. Smartphone's evolution allowed for developing system which rely on its sensors and processing capacities to present a solution which isn't dependant on specific hardware. With such a dependancy, the first condition that is imposed onto the systems is the compatibility with smartphones, i.e. the required sensors needs to exist on the generic hardware of smartphones. Once this barrier is surpassed, these generic system's are immediately faced with three fundamental questions which will define the architecture of the system and which can be seen on figure \ref{fig:choices}. The first question, Technology, defines the technology which is to be used in conjunction with the smartphone and consequently the way that location data is to be collected. There is a wide range of possibilities for this choice, be it BLE beacons, Wi-Fi Access points, LED lamps or just the microphone for sound collection, and each has an impact on the way that the system functions and on its performance. The second question is the location algorithm, which fundamentally depends on the target requirements of the system, if it's required to provide accurate location of a user or if a more descriptive location, such as the room in which the user is located, is enough. The third and last question is about which way will the location be described, with the existant possibilities having previously been presented.

@@

When taking into consideration every possible technology and existant algorithm that are capable of being applied on indoor location, we find ourselves with a wide variaty of solutions. This high number of variations has led to an increase on the number of existing surveys which attemp to gather make a collection of what has been achieved in the field. Looking at this opportunity one saw the possibility of attempting to create a generic platform capable of being utilized by any existing indoor location system. In order to implement and test the solution there was a need to decide on which technology to be used and as such one opted by the bluetooth low energy.

The Idealized generic architecture , which can be seen on figure \ref{fig:generic}, was created with the intention of providing a common ground for any systems which planned to have an indoor location based on a smartphone, which allows interoperability among those who make use of it, while presenting a way to structure the system so that the impact on the smartphone is reduced. If one analyses it according to the three points presented in figure \ref{fig:choices}, one can say that by having dedicated serves for the location computation and the location description instead of having it locally on the smartphone, interoperability is highly improved. This setup allows for systems to implement whichever algorithm they intent for as long as the server is capable of distinguishing which algorithm is to be used, responsability which is passed onto the smartphone. In scenarios of interoperability, the smartphone should be able to use any of its sensors to capture data, dealing with the technology component, and tag them before sending it for computation on the server. The location description is contained on the map server and it is up to the system to choose which one to utilise, be it self-created or owned by a third party such as google maps or openstreet maps. For as long as the smartphone is aware of the type of location that he is to receive, interoperability should be achieved.

@@

% Review

This paper is structured in a way that in \ref{sec:related} an overview of existent indoor location systems, generic and dedicated, is given, followed by an analyses on the existent location techniques and the ways to decribe a location. Section \ref{sec:architecture} overviews the architecture of the proposed generic system, while section \ref{sec:structure} presents the finalized system by analyzing the utilized technologies. Section \ref{sec:performance} analyzes the different aspects of the presented solution in terms of energetic efficiency, accuracy and response timings, finalizing in section \ref{sec:future} by overviewing what could be carried out in order to further develop the existing work and by concluding the paper.

@@

In this chapter one will first analyse the systems that layed the foundations of indoor location systems in section \ref{subsec:dedicated}, in order to study systems that were fully dependant on having specific hardware for obtaining user location. Section \ref{subsec:gentech} will present the technlogies that are compatible with modern day smartphones, analyse its pros and cons and mention existant projects based on them. The following section, \ref{subsec:technique} will describe the existent ways of obtaining the data that is required for infering a user's location as well as the techniques that make use of them. To conclude, on section \ref{subsec:desciption}, one will present the existent ways of describing locations through system that make use of them.

@@

In 1992 the Active badge system \cite{badge} was presented as an infrared solution which was capable of provided room-based position tracking, due to infrared signals being unable to travel through walls. This system imposed users to carry with them an ID card equiped with an IR LED which emitted a signal each fifteen seconds \cite{badge1, badge2}. The long signal frequency was due to the chance of collision of signals from multiple users in the same location while also having in mind energy cost which need to be kept as low as possible. One of the downsides was the decrease in accuracy as the user's location is only known at best to a 15-second window. The user's location was obtained  through the implementation of a network of sensors which act as receivers and then forward the obtained information to the master station which would take care of polling the sensors, processing data and then making it available to users. One of the biggest issues that were raised with this system was the privacy concerns raised by the clients due to its object tracking nature.

At the beggining of the twentieth century, two remarkable projects were presented, Bat system \cite{bat} and RADAR\cite{radar}. The first one, Bat Ultrasonic Location System, was a system capable of tracking various objects, each with a small wireless device called bat \cite{bat1, bat2}. A bat consisted of a radio transmitter with a unique identifier and was ment to be carried by personnel or attached to objects. For locating a bat, the base station periodically transmitted a signal with a single ID which caused the corresponding bat to emmit an ultrasound pulse. This pulse was then captured by an existing grid of receivers, which were placed on the ceiling of the areas to be covered, which recorded the time of arrival of the ultrasound. Since the speed of sound in the air is known, the system was capable of converting the time of flight of the ultrasound into distances, in order to obtain the bat's location.

The Radar system is a RF-based system to locate and track users inside buildings. The system is devided into two phases, an offline phase used for data collection and a real-time phase where a location is obtained \cite{radar1}. For data collection the mobile application and the base stations are synchronized, in order to be able to obtain time stamps and then the mobile user periodically emits UDP packets to the base stations (BS) . Each BS records the RF signal strength measurement together with the time stamp, in order to create a fingerprint of the map.  The real-time phase utilises the real-time user's packet associated RF signal strength, which is obtained at each bs and later on forwarded to the central computer where the computation is made, and examinates it in order to find the best fit for the current transmitter position.

The last remarkable project, Cricket, managed to tackled some of the problems presented in the last mentioned systems \cite{cricket1}. Each node in the cricket system is composed of a RF transceiver and hardware capable of generating and receiving ultrasonic signals. There are two types of nodes, beacons, which are fixed reference points and are attached to the ceiling or walls of the building, and listeners which are attached to the objects that need to be tracked. Each beacon periodically transmits a RF signal message containing beacon specific information, such as unique ID and beacon position. Whenever a RF signal is transmitted, an ultrasonic pulse is also emitted which enables listeners to measure their distance to the beacons by using the time difference of arrival times of the RF and ultrasonic signals. Each listener utilizes the RF signal's beacon information alongside the obtain distances to beacons to compute their space position and orientation. Crickets Architecture allowed to solve the user pribacy, decentralized and ease of deployment issues there were present in the remaining projects \cite{cricket}.

@@

Wi-Fi is a technology that can be used to estimate the location of a mobile user that resides inside the network. Nowadays Wi-Fi positioning systems have become the most widespread approach for indoor location systems since Wi-FI access points are readily available in many indoor environments and since all smartphones are equiped with a Wi-Fi antenna they can be located without the need of installing extra software or manipulating the hardware. One issue of Wi-Fi signals is that they suffer attenuation from static environment such as walls and movement of furniture and doors. The most widely location technique used is the radio signal strength indicator (RSSI), which suffers from severe multi-path effects leading to propagation model failures and as such inaccuracy in distance measurement. With these problems in mind a technique called RSSI-based fingerprinting is often used in order to improve performance, the same technique used in the Radar system.

Infrared (IR) systems are one of the most common position system that utilise wireless technology that has been used to track objects or people. IR wavelengths are invisible to the human eye under most circumstances, making this technology less intrusive than those which are visible. This technology is completely available in all smartphones since it makes use of the camera for data input, requiring only line-of-sight communication between receiver and transmitter, preferably without interference from strong light sources. One example is Epsilon \cite{epsilon}, which makes use of Light-emitting Diode (LED) and light sensors. Epsilon presents a solution which is easy to use in today's world due to the availability of LED lighting in indoor environments, with the only requirement being the adaption of the existing infrastructure to the required by the system. In order for the system to work, it required the mobile user to carry a device with an incorporated light sensor, which is the case of pretty much any existent smartphone.

Bluetooth is a wireless technology that was created in 1994 with the objective of replacing cables connecting fixed or portable devices. The Bluetooth Low Energy protocol was introduced with the Bluetooth Core Specification version 4 (also called Bluetooth Smart) \cite{BLECore} and it standed out for its lower power consuption, lower complexity and lower cost, while allowing for  device discovery, connection establishment and connection mechanisms.

@@

When looking at what's possible to achieve using the BLE technology there is the example of Apple's creation iBeacon \cite{ibeacon} which was presented in 2013 with the porpuse of implementing proximity sensing systems. The device is capable of playing on the broadcaster role and as such its objective is to send nearby compatible receivers certain information. Some examples of application are to track customers or trigger location-based actions on devices such as push notifications or checking in on social media, with pratical cases such as the usage of iBeacons by McDonalds to offer special offers to their customers in their fast-food stores. An indoor location system utilizing this technology was presented by Jingjing Yang et al \cite{ibeacon1}, where these devices were used to indicate a pacient of his whereabouts through the proximity sensing proprieties and this information was later transfered over to a server in order to give clients a variaty of different services, from pacient counting, to nearby department's information and offer indoor guidance to the nearest available bed.

When utilizing BLE for indoor location the usual metric used to calculate distances is the RSSI. This metric withint the context of bluetooth brings to surface several issues such as the fact that RSSI as a metric is very accurate only when the target is within a meter of the beacon, since the value decreases as the inverse of the square of the distance to the beacon . As such when developing solutions for indoor location that require system with high accuracy capable of tracking moving objects, the usage of RSSI can't be utilized without further work. Faragher et al \cite{bleacc} tackled one of the techniques used to improve BLE system's accuracy, fingerprinting, by verifying the effects caused by the device deployment density within the required location. This experiment also puts into evidence one of the downsides of the bluetooth technology being that its scalability is low, besides requiring higher density in order to increase accuracy, due to their low range any need to increase coverage leads to increased costs.

Zonith \cite{zonith} introduced a bluetooth based location system with the objective of tracking the position of workers in dangerous environments. Any device registered in the zonith implemented network would be continuously tracked and accounted for in each of the system's functionalities such as, sounding an alarm whenever a lone worker doesn't move or responde within a time interval (Lone worker protection) or providing a quick an precise location of any worker that has requested for help. This system's installation requires planing of the best locations to place the beacons and number required of beacons in order to be able to provide enough courage and make sure the system provides the required quality.

@@

The location of a user can be obtained through several different techniques, each possibility leads to different levels of accuracy and hardware and computacional costs.
The first technique would be proximity detection which has Cell of Origin (CoO) as its most famous technique and allows for room-based detection by assuming that the mobile target is at the cell with the strongest receiving power. This technique is widely used by system using RFID, Bluetooth and Infrared and has low deployment costs as the number of beacons required is small, at least one per room. 
The following technique would be triangulation which uses the geometric properties of the triangle to obtain a position. The metrics utilized to obtain the values used in the position's calculation can be angle-based, which have high implementation costs due to the required antenna's complexity; time-based, using Time-of-flight (ToF) or Round Trip Time (RTT) to obtain measures of distance from beacons; signal property based, using received signal strength indicator (RSSI), altough it's only possible of using this metric with radio signals; dead reckoning, which calculates the actual position utilizing the last determined position and incrementing based on estimate of the device's speed.

@@

In order to better understand the existent ways of describing locations it's better to look at existent solutions and the methods that they utilise. The most common method of describing location is through latitude and longitude, much like the gps does, as the biggest example that makes use of it is Google maps. Google maps, through its indoor maps platforms \cite{googlemaps} allows integration of indoor maps onto their google maps. Since google maps uses this type of location description, the indoor component of it follows the same routes. An indoor map on the platform is inserted onto its original geographical location and in the case of having multiple floors, it is possible to navigate through them.

Another method is to consider the indoor map as the reference and utilise a cartesion coordinate system, x and y, to represent a location. Meridian, an indoor location system developed by HP \cite{meridian} which presents functionalities such as route making and push notifications through its beacons, makes use of its platform for insertion of the indoor maps. As such there isn't the outdoor component that google indoor maps had,making it possible to use a (x,y) system relative to the building map.

Although Meridian already provides some extra level of detail in comparison with google indoor maps, there is another example that should be mentioned called OpenStreetMaps. Although it started much like google, providing global data, due to its openness, many indoor projects surfaced such as OpenLevelUP \cite{openlevel}. This project makes use of OpenStreetMaps' current indoor tagging scheme \cite{opentagging}, which is intented to describe in the most complete and simple way a building. This makes available the number of floors, the type of elemets (room, wall, corridor, etc) and its connectors, like doors and escalators or elevators, allowing to understand clearly the map that is being analysed.

@@

In order to understand how a system based on this generic architecture, figure \ref{fig:generic} has to be strutured, it is of relevance to analyse the system component's description and requirements that would allow it to be deployed. If one starts by analysing the beacons component, one can describe them as the reference points of the system which are in charge of providing nearby users with information relative to their surroundings. These beacons are always ready and available to the users and upon being contacted, they should inform the smartphone which has initiated the communication of the address of the beacon's associated location server. Since the beacon only communicates with the smartphone, it isn't required to have any other communication capacity other than that specific to its technology. The last requirement is that the beacon needs to be uniquely identified and have the address of its associated server, while being capable of transmitting that same information to the smartphone.

The smartphone service is the component that is in charge of communicating with both the beacons and the location server. This service is a software component of the smartphone application which is available to being called when a location is requested. Once a location is requested by the application, the service should communicate with nearby beacons through one of its sensors. A service is associated with one and only one technology, as such if the application is to support more than one technlogy, multiple services should be implemented. This condition is relevant since the required sensor depends on the chosen technology and because it allows the service to quickly associate the data that it is collecting with the technology, information which must be delivered to the location server. Once the beacon data has been collected, the same data is forwarded to the beacon's associated server, together with a tag representing the technology of the beacons. Upon obtaining a location, the service should forward the data to the application, which is responsible of later communicating with the map server. In terms of requirements, the service needs to have Wi-Fi or at least 3G available for communicating with both the location and maps server, and access to the sensor required to communicate with the nearby beacon, for BLE it would be the bluetooth antenna or for QR it would be the camera.

The Location server should have a database of all the devices associated to it, each with its exact location of the map. This location is dependent on the method for location description chosen, i.e. latitude/longitude or x/y. The server should also be able to take input from any of the supported technologies, depends on the beacons in the database, in order to output the user's location. From this description one can define the requirements of the location server as having in storage the beacons that are associated it as well as their location, being capable of communicating with the smartphone application and being in accordance with the map server, achieved through providing location description that is the same as that present in the maps server.

@@

The solution presented in this paper was made with the objective of creating a generic indoor location system capable of being implemented using any existant indoor system created. The generic system's architecture is presented in figure ~\ref{fig:generic} and it's divided in 4 parts: beacon, location server, map server and smartphone application.

The environment with beacons represents any form of element responsible for providing fixed reference points which are fundamental in calculating a user's position. A beacon needs to be compatible with the usage of a smartphone, i.e. the device's sensors need to be able to capture the data. Using the previously presented technologies as examples, these "beacons" could be in fact BLE beacons , Wi-Fi access points, LED lamps or even something as simple as sound, with the data capturing being made through the smartphone's available antennas, camera or microphone. The beacon represents a position in the indoor environments and as such it needs to be uniquely identified, be it through its characteristics or associated information. In addition to these informations, each beacon needs to have an associated location server and provide information about it to the smartphone application.

The location server is an external component where the information relative to the "beacons" utilised is stored. As such when a location request arrives from the smartphone, the received data can be translated onto physical locations which will later be used to compute the user's location. The algorithm used to obtain such location is entirely dependant on the system's creators. Once the location is obtain, it's sent back to the smartphone in addition to the location of the server where it should retrieve the associated map.

>>>>

\title{{ \normalsize Master Thesis} \\ 
GEFILOC: Generic Framework for Indoor Location Systems} 
\author{ 
Ricardo Martins {\tt ricardo.pachecomartins@gmail.com} 
Instituto Superior T\'{e}cnico} 
\date{\today}

@@

Nowadays there is a great demand for positioning systems and while the outdoor performance is amazing through the use of GPS, indoor positioning systems have been studied in order to also achieve such results. In this paper one studies the actual state of indoor systems, through an analysis on the usable technologies and techniques. A critical analysis is conducted through a modern perspective centered on location services through smartphones. Since there is a great number of different indoor location systems, one found a need to create a framework capable of hosting such systems in the form of a smartphone-centric architecture. This generic architecture attempted to achieve  interoperability  through the isolation of the main components of indoor systems. As such GEFILOC, a generic framework for indoor location systems, was created. GEFILOC would allow for systems that made use of it, to be capable of operating together on the same device without additional requirements. In order to test the proposed framework, a Bluetooth low energy-based system was implemented. This implementation allowed for analysing the energetic costs of the architecture and comprehending how to set up one or more systems for analysing the interoperability component.

@@

Outdoor positioning systems have greatly evolved in the past with the global positioning system (GPS) as its most important system. Since GPS is an outdoor position systems based on a network of satellite, when it's used for indoor location, it isn't capable of having the same performance. This is caused by the constraints introduced by indoor environments such as attenuation and reflection of electromagnetic waves upon collision with building walls and obstacles \cite{survey1}. As such there was a need to find reliable indoor systems that by nature would reduce the impact of some of the mentioned constraints.

In order to understand indoor position there is a need to understand the full scope of variables that comes to surface when moving from outdoor to indoor. When developing an indoor system there is a need to make sure that it can tackle the challenges such as small space dimension which reinforce the need for higher precision, a higher probability of non existent line of sight, influence of obstacles such as walls, furniture, moveable objects such as doors and human beings\cite{reviewtechniques, survey3}. All of the previously mentioned affect the way electromagnetic waves propagate in an indoor environment leading to problems related to severe multipath and reflection on existent surfaces \cite{surveywireless}. Besides propagation challenges, there are energy consumption, accuracy and deployment costs that play a critical role in deciding the viability of a proposed indoor location technique. Another interesting challenges that also surfaces is the means to describe a location. When dealing with outdoor location systems, the provided location is always characterised by two values, latitude and longitude, since the existent environment, i.e. the planet Earth, is always the same and as such the coordinates system can be relative to it. In indoor systems, there are many different types of surrounding environments and as such different ways of representing the data are necessary. If one thinks about providing indoor location on an office, the precise location isn't as relevant as just knowing the general location, i.e. a building specific description of the location in the form of building, floor and room. On other environments such as supermarkets or even in the previous example, where the objective is to provide more a precise location description, a cartesian coordinate system (x,y) is required.

With the evolution of mobile devices there has been a sizeable number of different technologies that can possibly be used in indoor location \cite{surveythesis,survey2} such as GPS-based technologies, using high sensitivity antenas to overcome GPS's indoor issues, RFID , Wireless LAN and Bluetooth among others, allowing even for hybrid systems which make use of more than one of the technologies mentioned above.

We are now in an era dominated by smartphones and as such they have became the central piece of several indoor location system. Smartphone's evolution allowed for developing system which rely on its sensors and processing capacities to present a solution that isn't dependent on specific hardware. With such a dependency, the first condition that is imposed onto the systems is the compatibility with smartphones, i.e. the required sensors needs to exist on the generic hardware of smartphones. Once this barrier is surpassed, these generic system's are immediately faced with three fundamental questions which will define the architecture of the system and which can be seen on figure \ref{fig:choices}. The first question, Technology, defines the technology that is to be used in conjunction with the smartphone and consequently the way that location data is to be collected. There is a wide range of possibilities for this choice, be it BLE beacons, Wi-Fi Access points, LED lamps or just the microphone for sound collection, and each has an impact on the way that the system functions and on its performance. The second question is the location algorithm, which fundamentally depends on the target requirements of the system, if it's required to provide accurate location of a user or if a more descriptive location, such as the room in which the user is located, is enough. The third and last question is about which way will the location be described, with the existent possibilities having previously been presented.

@@

When taking into consideration every possible technology and existent algorithm that are capable of being applied on indoor location, we find ourselves with a wide variety of solutions. This high number of variations has led to an increase in the number of existing surveys which attempt to gather make a collection of what has been achieved in the field. Looking at this opportunity one saw the possibility of attempting to create a generic platform capable of being used by any existing indoor location system. In order to implement and test the solution there was a need to decide on which technology to be used and as such one opted by the bluetooth low energy.

The idealised generic architecture, which can be seen on figure \ref{fig:generic}, was created with the intention of providing a common ground for any systems which planned to have an indoor location based on a smartphone. It would allow for interoperability among those who make use of it, while presenting a way to structure the system so that the impact on the smartphone is reduced. If one analyses it according to the three points presented in figure \ref{fig:choices}, one can say that by having dedicated servers for the location calculation and the location description instead of having it locally on the smartphone, interoperability is highly improved. This setup allows for systems to implement whichever algorithm they intent, for as long as the server is capable of distinguishing which algorithm is to be used. In scenarios of interoperability, the smartphone should be able to use any of its sensors to capture data, dealing with the technology component, and tag them before sending it for computation on the server. The location description is contained on the map server and it is up to the system to choose which one to use, be it self-created or owned by a third party such as google maps or openstreet maps. For as long as the smartphone is aware of the type of location that he is to receive, interoperability should be achieved.

@@

This paper is structured in a way that in section \ref{sec:related} an overview of existent indoor location systems, generic and dedicated, is given, followed by an analyses on the existent location techniques and the ways to describe a location. Section \ref{sec:system} is entirely dedicated to the GEFILOC. The section starts by describing the framework's requirements in subsection \ref{subsec:deployment}, followed by an overview of the architecture in subsection \ref{subsec:architecture}. Subsection \ref{sec:struture} presents the Bluetooth low energy implemented system, while subsection \ref{sec:eval} presents a critical analysis on the implemented system's energetic consumption.

@@

In this chapter one will first analyse the systems that laid the foundations of indoor location systems in section \ref{subsec:dedicated}, in order to study systems that were fully dependent on having specific hardware for obtaining user location. Section \ref{subsec:gentech} will present the technologies that are compatible with modern day smartphones, analyse its pros and cons and mention existent projects based on them. The following section, \ref{subsec:tecnique} will describe the existent ways of obtaining the data that is required for inferring a user's location as well as the techniques that make use of them. To conclude, on section \ref{subsec:description}, one will present the existent ways of describing locations through system that make use of them.

@@

In 1992 the Active badge system \cite{badge} was presented as an infrared solution which was capable of provided room-based position tracking, due to infrared signals being unable to travel through walls. This system imposed users to carry with them an ID card equipped with an IR LED which emitted a signal each fifteen seconds \cite{badge1, badge2}. The long signal frequency was due to the chance of collision of signals from multiple users in the same location while also keeping in mind energy cost, which need to be kept as low as possible. One of the downsides was the decrease in accuracy as the user's location is only known at best to a 15-second window. The user's location was obtained  through the implementation of a network of sensors, which acted as receivers and forwarded the obtained information to the master station. The master station was responsible for polling the sensors, processing data and then making it available to users. One of the biggest issues that were raised with this system was the privacy concerns raised by the clients due to its object tracking nature.

At the beginning of the twentieth century, two remarkable projects were presented, Bat system \cite{bat} and RADAR\cite{radar}. The first one, Bat Ultrasonic Location System, was a system capable of tracking various objects, each with a small wireless device called bat \cite{bat1, bat2}. A bat consisted of a radio transmitter with a unique identifier and was meant to be carried by personnel or attached to objects. For locating a bat, the base station periodically transmitted a signal with a single ID which caused the corresponding bat to emit an ultrasound pulse. This pulse was then captured by an existing grid of receivers, which were placed on the ceiling of the areas to be covered, which recorded the time of arrival of the ultrasound. Since the speed of sound in the air is known, the system was capable of converting the time of flight of the ultrasound into distances, in order to obtain the bat's location.

The Radar system is a RF-based system to locate and track users inside buildings. The system is divided into two phases, an offline phase used for data collection and a real-time phase where a location is obtained \cite{radar1}. For data collection the mobile application and the base stations are synchronised, in order to be able to obtain time stamps and then the mobile user periodically emits UDP packets to the base stations (BS) . Each BS recorded the RF signal strength measurement together with the time stamp, in order to create a fingerprint of the map.  The real-time phase utilises the real-time user's packet associated RF signal strength, that is obtained at each base station. This information is then forwarded to the central computer where the computation is made.

The last remarkable project, Cricket, managed to tackled some of the problems presented in the last mentioned systems \cite{cricket1}. Each node in the cricket system is composed of a RF transceiver and hardware capable of generating and receiving ultrasonic signals. There are two types of nodes, beacons, which are fixed reference points and are attached to the ceiling or walls of the building, and listeners which are attached to the objects that need to be tracked. Each beacon periodically transmits a RF signal message containing beacon specific information, such as unique ID and beacon position. Whenever a RF signal is transmitted, an ultrasonic pulse is also emitted which enables listeners to measure their distance to the beacons by using the time difference of arrival times of the RF and ultrasonic signals. Each listener utilizes the RF signal's beacon information alongside the obtain distances to beacons to compute their space position and orientation. Cricket's architecture allowed to solve the user privacy, decentralised and ease of deployment issues there were present in the remaining projects \cite{cricket}.

@@

Wi-Fi is a technology that can be used to estimate the location of a mobile user that resides inside the network. Nowadays Wi-Fi positioning systems have became the most widespread approach for indoor location systems since its access points are readily available in many indoor environments and since all smartphones are equipped with a Wi-Fi antenna they can be located without the need of installing extra software or manipulating the hardware. One issue of Wi-Fi signals is that they suffer attenuation from static environment such as walls and movement of furniture and doors. The most widely location technique used is the radio signal strength indicator (RSSI), which suffers from severe multi-path effects leading to propagation model failures and as such inaccuracy in distance measurement. With these problems in mind a technique called RSSI-based fingerprinting is often used in order to improve performance, the same technique used in the Radar system.

Infrared (IR) systems are mostly used for tracking objects or people. IR wavelengths are invisible to the human eye under most circumstances, making this technology less intrusive than those which are visible. This technology is completely available in all smartphones since it makes use of the camera for data input, requiring only line-of-sight communication between receiver and transmitter, preferably without interference from strong light sources. One example of an IR system is Epsilon \cite{epsilon}, which makes use of Light-emitting Diode (LED) and light sensors. Epsilon presents a solution which is easy to use in today's world due to the availability of LED lighting in indoor environments, with the only requirement being the adaptation of the existing infrastructure to the required by the system. In order for the system to work, it required the mobile user to carry a device with an incorporated light sensor, which is the case of pretty much any existent smartphone.

Bluetooth is a wireless technology that was created in 1994 with the objective of replacing cables connecting fixed or portable devices. The Bluetooth Low Energy protocol was introduced with the Bluetooth Core Specification version 4 (also called Bluetooth Smart) \cite{BLECore} and it stood out for its lower power consumption, lower complexity and lower cost, while allowing for  device discovery, connection establishment and connection mechanisms.

@@

When looking at what's possible to achieve using the BLE technology there is the example of Apple's creation iBeacon \cite{ibeacon} which was presented in 2013 with the purpose of implementing proximity sensing systems. The device is capable of being a broadcaster, whose objective is to send information to nearby compatible receivers. The iBeacon can be used to track customers or trigger location-based actions on devices such as push notifications or checking in on social media. As an example, McDonalds used iBeacons to offer special offers to their customers in their fast-food stores. An indoor location system using this technology was presented by Jingjing Yang et al \cite{ibeacon1}, where these devices were used to indicate a patient of his whereabouts through the proximity sensing properties. The user's location was later transferred to a server in order to give clients a variety of different services, from patient counting, to nearby department's information and offer indoor guidance to the nearest available bed.

When employing BLE for indoor location, the usual metric used to calculate distances is the RSSI. This metric within the context of Bluetooth brings to surface several issues such as the fact that RSSI as a metric is very accurate only when the target is within a meter of the beacon, since the value decreases as the inverse of the square of the distance to the beacon. As such when developing solutions for indoor location that require systems with high accuracy, capable of tracking moving objects, the usage of RSSI isn't viable without further work. Faragher et al \cite{bleacc} tackled one of the techniques used to improve BLE system's accuracy, fingerprinting, by verifying the effects caused by the device deployment density within the required location. This experiment also puts into evidence one of the downsides of the Bluetooth technology being that its scalability is low. Due to their low range, in order to increase accuracy, an higher density is required .

Zonith \cite{zonith} introduced a Bluetooth based location system with the objective of tracking the position of workers in dangerous environments. Any device registered in the zonith implemented network would be continuously tracked and accounted for in each of the system's functionalities. The system's functionalities involve sounding an alarm whenever a lone worker doesn't move or respond within a time interval (Lone worker protection) or providing a quick a precise location of any worker that has requested for help. This system's deployment requires a study on the best locations for placing the beacons and on the number of required beacons, so that it can provide enough coverage and assure the system's quality.

@@

The location of a user can be obtained through several different techniques, with each possibility leading to different levels of accuracy and hardware and computational costs. 
The first technique would be proximity detection which has Cell of Origin (CoO) as its most famous technique and allows for room-based detection by assuming that the mobile target is at the cell with the strongest receiving power. This technique is widely used by system using RFID, Bluetooth and Infrared and has low deployment costs as the number of beacons required is small, at least one per room.  
The following technique would be triangulation which uses the geometric properties of the triangle to obtain a position. The metrics used in the position's calculation can be angle-based, which have high implementation costs due to the required antenna's complexity; time-based, using Time-of-flight (ToF) or Round Trip Time (RTT) to obtain measures of distance from beacons; signal property based, using received signal strength indicator (RSSI), although it's only possible of using this metric with radio signals; dead reckoning, which calculates the actual position through the last determined position and an estimate of the device's speed.

@@

In order to better understand the existent ways of describing locations it's better to look at existent solutions and the methods that they employ. The most common method of describing location is through latitude and longitude, much like the GPS does, a method used by Google maps. Google maps, through its indoor maps platforms \cite{googlemaps} allows integration of indoor maps onto their maps. Since Google Maps uses this type of location description, its indoor component follows the same format. An indoor map on the platform is inserted into its original geographical location and in the case of having multiple floors, it is possible to navigate through them.

Another method is to consider the indoor map as the reference and employ a cartesian coordinate system, x and y, to represent a location. Meridian, an indoor location system developed by HP \cite{meridian}, which presents functionalities such as route making and push notifications through its beacons, makes use of its platform for insertion of the indoor maps. As such there isn't the outdoor component that google indoor maps has, making it possible to use a (x,y) system relative to the building map.

Although Meridian already provides some extra level of detail in comparison with google indoor maps, there is another example that should be mentioned, OpenStreetMaps. Although it started much like google, providing global data, due to its openness, many indoor projects surfaced such as OpenLevelUP \cite{openlevel}. This project makes use of OpenStreetMaps' current indoor tagging scheme \cite{opentagging}, which is intended to describe in the most complete and simple way a building. This makes available the number of floors, the type of elements (room, wall, corridor, etc) and its connectors, like doors and escalators or elevators, allowing to understand clearly the map that is being analysed.

@@

In order to understand how a system based on this generic architecture, which can be seen on figure \ref{fig:generic}, has to be structured, it is of relevance to analyse the system component's description and requirements that would allow it to be deployed. If one starts by analysing the beacons component, one can describe them as the reference points of the system, which are in charge of providing nearby users with information relative to their surroundings. These beacons are always ready and available to the users and upon being contacted, they should inform the smartphone which has initiated the communication of the address of the beacon's associated location server. Since the beacon only communicates with the smartphone, it isn't required to have any other communication capacity other than that specific to its technology. The last requirement is that the beacon needs to be uniquely identified and have the address of its associated server, while being capable of transmitting that same information to the smartphone.

The smartphone service is the component that is in charge of communicating with both the beacons and the location server. This service is a software component of the smartphone application which is available to being called when a location is requested. Once a location is requested by the application, the service should communicate with nearby beacons through one of its sensors. A service is associated with one and only one technology, making it so that if the application is to support more than one technology, multiple services should be implemented. This condition is relevant since the required sensor is dependent on the chosen technology, and because it allows the service to quickly associate the data that it is collecting with the technology. Once the beacon data has been collected, it's forwarded to the beacon's associated server, together with a tag representing the technology of the beacons. Upon obtaining a location, the service should pass the data to the application, which is responsible of later communicating with the map server. In terms of requirements, the service needs to have Wi-Fi or at least 3G available for communicating with both the location and maps server, as well as access to the sensor required to communicate with the nearby beacon. In the case of BLE, the sensor would be the Bluetooth antenna, while for QR it would be the camera.

The Location server should have a database of all the devices associated to it, each with its exact location of the map. This location is dependent on the method for location description chosen, i.e. latitude/longitude or x/y. The server needs to be capable of taking input from any of the supported technologies, in order to output the user's location. From this description one can define the requirements of the location server as: having in storage the beacons that are associated it as well as their location; being capable of communicating with the smartphone application; and being in accordance with the map server, achieved through providing location description that is the same as that present in the maps server.

@@

The solution presented in this paper was made with the objective of creating a generic indoor location system capable of being implemented using any existent indoor system created. The generic system's architecture is presented in figure ~\ref{fig:generic} and it's divided in 4 parts: beacon, location server, map server and smartphone application.

The environment with beacons represents any form of element responsible for providing fixed reference points which are fundamental in calculating a user's position. A beacon needs to be compatible with the usage of a smartphone, i.e. the device's sensors need to be able to capture the data. Using the previously presented technologies as examples, these "beacons" could be in fact BLE beacons , Wi-Fi access points, LED lamps or even something as simple as sound, with the data collection being made through the smartphone's available antennas, camera or microphone. The beacon represents a position in the indoor environments and it is uniquely identified, be it through its characteristics or associated information. In addition to these informations, each beacon needs to have an associated location server and provide information about it to the smartphone application.

The location server is an external component where the information relative to the beacons is stored. As such when a location request arrives from the smartphone, the received data can be translated into physical locations, which will later be used to calculate the user's location. Once the location is obtained, it's sent back to the smartphone in addition to the location of the server where it should retrieve the associated map.

The map server represents the architectural block responsible for providing the maps associated to the location of the user. The used map representation is up to the system for as long as it is in accordance with the remaining parts of the system. This means that the location provided by the location server needs to be representable on the maps provided and capable of being comprehended by the smartphone application.
